Verification of the CVM algorithm with a New Recursive Analysis Technique

### 2025-02-21 paper/sec/invariants.tex
Paper: More details on the derivation of the upper tail bound.

Of course, the exponential function in $M(t)$ is convex, so this is not yet the full story.
However, we can instead try to derive tail bounds for the random variable $\indicat(p \geq q) p^{-1} |\chi|$.
We end up with the condition that $h$ needs to be non-negative and concave only on $[0;q^{-1}]$.
This then allows us to use approximate the exponential function from above with an affine function $h$ on the range $[0;q^{-1}]$, which yields tail bounds for $p^{-1} |\chi|$ under the condition $p \geq q$. As an example, the upper tail bound can be derived as follows:
\begin{align*}
  \prob( p^{-1} |\chi| \geq (1+\varepsilon) |A| \wedge p \geq q) & \leq 
  \prob( I(p \geq q) p^{-1} |\chi| \geq (1+\varepsilon) |A|) \\
  & \leq \exp(-t(1+\varepsilon) |A|) M(t) \\ % = exp(-t(1+\varepsilon) |A|) \expect \left[ \prod_{s \in A} \exp( t p^{-1} X_s) \right]
  & \leq \exp(-t(1+\varepsilon)|A|) \expect \left[ I(p \geq q) \prod_{s \in A} h(p^{-1} X_s) \right] \\
  & \leq \exp(-t(1+\varepsilon)|A|) h(1)^{|A|} \leq \exp(-\frac{n}{12} \varepsilon^2)
\end{align*}
where we choose $h(x) = 1+qx (\exp(t/q)-1)$.
For the last step, we have to find the $t$ that produces the required bound.%
\footnote{We use $t = q \ln(1+\varepsilon)$ which is not the real optimum, but better for algebraic evaluation.} 
Note that $h$ is affine and it can be easily checked%
\footnote{Because the exponential function is convex and $h$ is affine, we only have to check the end points: $0, q^{-1}$.}
that it is an upper approximation of $\exp(tx)$ for $x \in [0;q^{-1}]$.
To use these bounds, we also have to separately estimate $\prob(p < q)$.
We use a similar strategy, as in the original proof by Chakraborty et al.~\cite{chakraborty2022}, where they use $q = \frac{n}{4 |A|}$.

>>>>

Of course, the exponential function in $M(t)$ is convex, so this is not yet the full story.
However, we can instead try to derive tail bounds for the random variable $\indicat(p \geq q) p^{-1} |\chi|$.
This leads to a similar inequality:
\begin{align}
  \expect \left[ \prod_{s \in A} I(p \geq q) h(p^{-1} X_s) \right] \leq h(1)^{|A|} \label[ineq]{i:func_invariant_capped}
\end{align}
with the new condition that $h$ needs to be non-negative and concave only on $[0;q^{-1}]$.
This then allows us to use approximate the exponential function from above with an affine function $h$ on the range $[0;q^{-1}]$, which yields tail bounds for $p^{-1} |\chi|$ under the condition $p \geq q$. As an example, the upper tail bound can be derived as follows:
\begin{eqnarray*}
  \prob( p^{-1} |\chi| \geq (1+\varepsilon) |A| \wedge p \geq q) & \leq &
  \prob( I(p \geq q) p^{-1} |\chi| \geq (1+\varepsilon) |A|) \\
  & \underset{\textrm{Markov}}{\leq} & e^{-t(1+\varepsilon) |A|} \expect \left[ \prod_{s \in A} I(p \geq q) \exp( t p^{-1} X_s) \right] \\
  & \leq & e^{-t(1+\varepsilon) |A|} \expect \left[ \prod_{s \in A} I(p \geq q) h(p^{-1} X_s) \right] \\
  & \underset{\textrm{Ineq.~\ref{i:func_invariant_capped}}}{\leq} & e^{-t(1+\varepsilon) |A|} h(1)^{|A|} \\
  & \underset{\textrm{By calculus}}{\leq} & \exp\left(-\frac{n}{12} \varepsilon^2\right)
\end{eqnarray*}
where we choose $h(x) = 1+qx (e^{t/q}-1)$.
Note that $h$ is affine and it can be easily checked%
\footnote{Because the exponential function is convex and $h$ is affine, we only have to check the end points: $0, q^{-1}$.}
that it is an upper approximation of $\exp(tx)$ for $x \in [0;q^{-1}]$.
For the last step, we have to find the $t$ that produces the required bound.%
\footnote{We use $t = q \ln(1+\varepsilon)$ which is not the real optimum, but better for algebraic evaluation.}
To use these bounds, we also have to separately estimate $\prob(p < q)$.
We use a similar strategy, as in the original proof by Chakraborty et al.~\cite{chakraborty2022}, where they use $q = \frac{n}{4 |A|}$.
